### 2015-06-28 doc/thesis/typedlua.tex
Updating strict versus default

Figure \ref{fig:types} presents the concrete syntax of
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and unions of tuple types.

@@

Figure \ref{fig:functions} shows that Typed Lua uses second-level
types in the definition of function types.
A second-level type is either the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
The type void \texttt{()} represents the type of the empty tuple.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuple types appear in the return type of function types to
represent overloading on the return type.
We will explain union types in more detail in the next section.
We can use only one first-level type \texttt{t} in the return type
because it is syntactic sugar to the tuple type \texttt{(t)}.

Typed Lua provides two operation modes: the \emph{default} mode and
the \emph{strict} mode.
In the default mode, the compiler adds a variadic tail to the
type of the parameter list and to the return type if the
programmer does not specify one.
The default mode has this behavior to preserve the semantics
of function calls in Lua, that is, it discards extra arguments and
uses \texttt{nil} to replace missing arguments.
In the strict mode, the compiler does not add a variadic tail
to the type of the parameter list and to the return type.
Instead, it preserves the type defined by the programmer to catch
arity mismatch.

@@

In the strict mode, \texttt{sum} has type
\texttt{(integer, integer) -> (integer)}.
The compiler does not add \texttt{value*} to the type of the parameter
list to catch arity mismatch.
For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
because we are passing an extra argument \texttt{3} to \texttt{sum}.

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The function call \texttt{v()} compiles without any compile-time warnings in
the default and strict modes because \texttt{(value*)} and \texttt{()}
are consistent with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
Calling \texttt{v(...)} compiles without any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

The type of the function \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The function call \texttt{message("Typed Lua")} compiles without
any compile-time warnings in both modes, because the argument list
\texttt{("Typed Lua")} has type \texttt{("Typed Lua", nil*)} that
is consistent with the input type \texttt{(string, string|nil, value*)}
in the default mode, and that is also consistent with the input type
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.
In Section \ref{sec:assignment} we will formalize this behavior.

>>>>

Figure \ref{fig:types} presents the concrete syntax of
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include vararg types, tuple types, and unions of tuple types.

@@

Figure \ref{fig:functions} shows that Typed Lua uses second-level
types in the definition of function types.
A second-level type is either a tuple of first-level types optionally
ending in a variadic type or a union of these tuples.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
The empty tuple \texttt{()} is syntactic sugar to \texttt{(nil*)},
as we will show that Typed Lua always ends a tuple type in a variadic type.
Union of tuple types appear in the return type of function types to
represent overloading on the return type.
We will explain union types in more detail in the next section.
We can use only one first-level type \texttt{t} in the return type
because it is syntactic sugar to the tuple type \texttt{(t)}.

Typed Lua provides two operation modes: the \emph{default} mode and
the \emph{strict} mode.
In the default mode, the compiler adds \texttt{value*} to the type
of the parameter list and \texttt{nil*} to the return type when
the programmer does not specify a variadic tail. 
It has this behavior to preserve the semantics
of function calls in Lua, that is, it uses \texttt{value*} to discard
extra arguments and \texttt{nil*} to replace missing arguments.
In the strict mode, the compiler adds \texttt{nil*} to both
parts of the function type.
Even though it still uses \texttt{nil*} to replace missing values,
it also uses \texttt{nil*} instead of \texttt{value*} to catch arity mismatch.

@@

In the strict mode, \texttt{sum} has type
\texttt{(integer, integer, nil*) -> (integer, nil*)}.
The compiler adds \texttt{nil*} to the type of the parameter
list to catch arity mismatch.
For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
because we are passing an extra argument \texttt{3} to \texttt{sum}.

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The function call \texttt{v()} compiles without any compile-time warnings in
both modes because the type of the argument list is \texttt{(nil*)},
which is consistent with \texttt{(string*)}.
The call \texttt{v("hello", "world")} also compiles without any warnings
in both modes because the type of the argument list is \texttt{("hello", "world", nil*)},
which is consistent with \texttt{(string*)}.
Calling \texttt{v(...)} compiles without any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

The type of the function \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)} in the
default mode, and
\texttt{(string, string|nil, nil*) -> (string, nil*)}
in the strict mode.
The call \texttt{message("Typed Lua")} compiles without
any compile-time warnings in both modes, because the argument list
\texttt{("Typed Lua")} has type \texttt{("Typed Lua", nil*)} that
is consistent with the input type \texttt{(string, string|nil, value*)}
in the default mode, and that is also consistent with the input type
\texttt{(string, string|nil, nil*)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.
In Section \ref{sec:assignment} we will formalize this behavior.
